---
title: "CSR Naloxone Standing Order Project - Draft Graphics"
author: 'STAT 245, Fall 2020 (Group Members: Nana Ama Baidoo, Alex Visser, Joshua Ridder, Joseph Jinn)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    fig_height: 8
    fig_width: 13
  html_document:
    fig_height: 8
    fig_width: 13
classoption: landscape
---

```{r, setup, include = FALSE}
# load packages that are going to be used
require(tidyverse)
require(ggformula)
require(mosaic)
require(fastR2)
require(s245)
require(pander)
require(DHARMa)
require(glmmTMB)
require(MuMIn)
require(car)
require(dplyr) # SQL syntax ftw.
require(readr)
require(ggeffects)
# require(reticulate) # Utilize Python programming language.

# Some customization.  You can alter or delete as desired (if you know what you are doing).

theme_set(theme_minimal(base_size=12))     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  echo = TRUE,      # for homework, always show R code (this is the default)
  error = TRUE, # do not interrupt generation in case of errors,
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small",   # slightly smaller font for code
  message = FALSE, warning = FALSE) # don't print warnings or messages in compiled document. So you MUST check them in RStudio!
```

Note to self: Run "remotes::install_github('ProjectMOSAIC/ggformula')" to install development version of ggformula.

```{r}
# sessionInfo()
options(max.print = 6000) # Ensure we can print entire summary.
# remotes::install_github('ProjectMOSAIC/ggformula')
```

# Initial import and preparation.

Import data.

```{r}
data <- read_csv2("datasets\\joseph_jinn_queries\\CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneAllCausesOfDeathDates9decimalsPer1k_TIDY.csv")
```

```{r}
glimpse(data)
```

### Convert columns to appropriate data types.

```{r}
data <- data %>% mutate_if(is.character, as.factor)
data <- data %>% mutate(county_code = as.factor(county_code))
data <- data %>% mutate(year = as.factor(year))
data <- data %>% mutate(year_code = as.factor(year_code))
data <- data %>% mutate(percent_of_total_deaths = as.numeric(gsub("%", "", percent_of_total_deaths)))
data <- data %>% mutate(deaths = as.integer(deaths))
data <- data %>% mutate(population = as.integer(population))
```

```{r}
glimpse(data)
```

```{r}
head(data)
```

___


# EDA - Exploratory Data Analysis (will correct labels, etc., in the future)

**These are initial EDA graphs for the entire dataset (drug, alcohol, and all non-drug/alchol causes of death).**

```{r}
gf_bar(~gender, data = data, alpha = 0.25, color = "#087EC6",
       xlab="Gender",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + labs(fill = "Vehicle Searched?") + gf_jitter()
```

```{r}
gf_bar(~race, data = data, alpha = 0.25, color = "#087EC6",
       xlab="Race",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~year, data = data, alpha = 0.25, color = "#087EC6",
       xlab="Year",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
)
```

```{r}
gf_bar(~ten_year_age_groups, data = data, alpha = 0.25, color = "#087EC6",
       xlab="10 Year Age Groups",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~county, data = data, alpha = 0.25, color = "#087EC6",
       xlab="County",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~cause_of_death, data = data, alpha = 0.25, color = "#087EC6",
       xlab="Cause of Death",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
)
```

```{r}
gf_point(deaths ~ population, data = data, alpha = 0.1, color = "#087EC6",
         xlab="Deaths",
         ylab="Population",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

```{r}
gf_point(deaths ~ crude_rate_added, data = data, alpha = 0.1, color = "#087EC6",
         xlab="Crude Rate Added",
         ylab="Deaths",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

```{r}
gf_point(deaths ~ percent_of_total_deaths, data = data, alpha = 0.25, color = "#087EC6",
         xlab="Percent of Total Deaths",
         ylab="Deaths",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

Quick and dirty EDA.

___


**These are initial EDA graphs for the subset of the dataset with ONLY drug-related deaths.**

```{r}
drug_deaths_only_data <- data %>% filter(data$cause_of_death == "drug")
```

```{r}
glimpse(drug_deaths_only_data)
```

```{r}
gf_bar(~gender, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="Gender",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + labs(fill = "Vehicle Searched?") + gf_jitter()
```

```{r}
gf_bar(~race, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="Raceh",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~year, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="Year",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
)
```

```{r}
gf_bar(~ten_year_age_groups, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="10 Year Age Groups",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~county, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="County",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
) + coord_flip()
```

```{r}
gf_bar(~cause_of_death, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
       xlab="Cause of Death",
       ylab="Number of Observations",
       title="CDC Wonder Underlying Cause of Death Dataset Queries",
       subtitle = "",
       caption = ""
)
```

```{r}
gf_point(deaths ~ population, data = drug_deaths_only_data, alpha = 0.1, color = "#087EC6",
         xlab="Deaths",
         ylab="Population",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

```{r}
gf_point(deaths ~ crude_rate_added, data = drug_deaths_only_data, alpha = 0.1, color = "#087EC6",
         xlab="Crude Rate Added",
         ylab="Deaths",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

```{r}
gf_point(deaths ~ percent_of_total_deaths, data = drug_deaths_only_data, alpha = 0.25, color = "#087EC6",
         xlab="Percent of Total Deaths",
         ylab="Deaths",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
)
```

___
___


```{r}
drug_data <- data %>% filter(data$cause_of_death == "drug")
```

And we should probably scale our predictors try to resolve some NA issues with coefficients.

```{r}
min(drug_data$deaths)
max(drug_data$deaths)
min(drug_data$population)
max(drug_data$population)
min(drug_data$crude_rate_added)
max(drug_data$crude_rate_added)
min(drug_data$percent_of_total_deaths)
max(drug_data$percent_of_total_deaths)
```

```{r}
drug_data <- drug_data %>%
  mutate(scaled_crude_rate_added = scale(crude_rate_added),
         scaled_percent_of_total_deaths = scale(percent_of_total_deaths),
         scaled_population = scale(population)
  )
```

```{r}
glimpse(drug_data)
```

# Model Assessment Plots (for the 1st research question).
```{r}
model_nbinom2 <- glmmTMB(deaths ~ gender + race + year + ten_year_age_groups + 
                           scaled_population + scaled_percent_of_total_deaths,
                         data = drug_data, 
                         family = nbinom2(link = "log"))
```

```{r}
summary(model_nbinom2)
```

```{r}
overdisp_fun(model_nbinom2)
```

Over-dispersion looks good as it's below 2-3 units.

```{r}
gf_acf(~model_nbinom2)
```

```{r}
gf_acf(~model_nbinom2) %>% gf_lims(y = c(-0.2, 0.2), x = c(0, 30))
```

And we have some issues with independence of residuals at two of the lags.  This doesn't look terrible though.


```{r}
model_sim <- simulateResiduals(model_nbinom2, n = 1000)
```

```{r}
gf_point(model_sim$scaledResiduals ~ fitted(model_nbinom2)) %>%
  gf_labs(x = "Predicted # of Deaths",
          y = 'Scaled Residuals')
```

And this definitely isn't a uniform distribution.  I'm going to guess we have issues with linearity and possibly mean-variance.

Could also check log(response variable) vs. predictors and scaled residuals vs. predictors but at this point it's clear this model fails.

___

# Prediction plots (for the 1st research question).

```{r}
ggpredict(model_nbinom2)
```

```{r}
# t(get_fixed(logit_model_no_offset)) %>% pander()
```

```{r}
# get_fixed(model_nbinom2)
```

The other two functions to obtain fixed values for predictors fails. So, going with ggpredict().

```{r}
glimpse(drug_data)
```

Prediction plot for "gender":

And the long-winded manual way...

```{r}
# note: "chr" variables have unique() values, "fct" variables have level()s
gender_prediction_data <- expand.grid(
  # char_variable = pull(police_stop, char_variable) %>% unique(),
  # quantitative_variable = seq(from = 0, to = 10000, by = 100),
  # categorical_variable = pull(police_stop, categorical_variable) %>% factor() %>% levels(),
  gender = pull(drug_data, gender) %>% factor() %>% levels(),
  race = "Black or African American",
  year = "1999",
  ten_year_age_groups = "15-24 years",
  scaled_population = -0.00,
  scaled_percent_of_total_deaths = 0.00)

# compute predictions with SEs
gender_predictions <- predict(model_nbinom2, 
                              newdata = gender_prediction_data,
                              type = 'response', se.fit = TRUE)

glimpse(gender_predictions)
```

```{r}
# add predicted values and CIs to the hypothetical dataset
gender_prediction_data <- gender_prediction_data %>%
  mutate(predictions = gender_predictions$fit,
         CI_low = gender_predictions$fit - 1.96 * gender_predictions$se.fit,
         CI_hi = gender_predictions$fit + 1.96 * gender_predictions$se.fit)
```

```{r}
gf_point(predictions ~ gender, data = gender_prediction_data,
         alpha = 1.0, color = "black",
         ylab="Predicted (fitted) Values of Deaths",
         xlab="Gender",
         title="CDC Wonder Data - Prediction Plot",
         subtitle = "",
         caption = ""
) %>%
  # gf_ribbon(CI_low + CI_hi ~ single_year_ages_factor) %>%
  gf_errorbar(CI_low + CI_hi ~ gender) %>% gf_lims(y = c(15, 25))
```

Prediction plot for "race":

```{r}
ggpredict(model_nbinom2, 
          terms = 'race',
          type = 'fixed') %>% plot()
```

Prediction plot for "year":

```{r}
ggpredict(model_nbinom2, 
          terms = 'year',
          type = 'fixed') %>% plot()
```

Prediction plot for "ten_year_age_groups":

```{r}
ggpredict(model_nbinom2, 
          terms = 'ten_year_age_groups',
          type = 'fixed') %>% plot()
```

Prediction plot for "population":

```{r}
ggpredict(model_nbinom2, 
          terms = 'scaled_population',
          type = 'fixed') %>% plot()
```
Prediction plot for "scaled_percent_of_total_deaths":

```{r}
ggpredict(model_nbinom2, 
          terms = 'scaled_percent_of_total_deaths',
          type = 'fixed') %>% plot()
```

___

# Model Assessment Plots (for the 3rd research question).

Convert to binary response and store in new variable.  Then, reassign as "police_stop".
```{r}
all_data <- data %>% mutate(drug_death = case_when(
  cause_of_death == "drug" ~ "yes",
  TRUE ~ "no"))
```

Convert response to a factor.

```{r}
all_data <- all_data %>% mutate(drug_death = as.factor(drug_death))
```

```{r}
all_data <- all_data %>% mutate(drug_death = fct_relevel(drug_death, c("no", "yes")))
```

```{r}
levels(all_data$drug_death)
```

```{r}
glimpse(all_data)
```

```{r}
model_logit <- glm(drug_death ~ gender + race + year + ten_year_age_groups + 
                     population + deaths + percent_of_total_deaths + gender*race,
                   data = all_data, family = binomial(link = 'logit'), na.action = "na.fail")
```

```{r}
summary(model_logit)
```

According to Week 7 slides on Binary Regression...

* Response variable is logical (data can be proportion but not binary!)
* Linearity: logit(p) should have a linear relationship with each predictor variable. (A bit hard to check - can bin predictor and then plot scatter plot)
* Independence of Residuals: Same as usual.
* Mean-variance relationship: scaled residuals have uniform distribution and no trends.
* NO distributional assumptions about residuals.

___

Our response variable "drug_death" is logical (since I made it so).  So, this check passes with flying colors.

___

Auto-Correlation Plots:

```{r}
acf(resid(model_logit))
```

Ok, that fails miserably right from the start.

Attempt to fix...

```{r}
require(lme4)
```

```{r}
model_logit_fix_attempt <- glmer(drug_death ~ gender + race + year + ten_year_age_groups + 
                     population + deaths + percent_of_total_deaths + gender*race + (1 | county),
                   data = all_data, family = binomial(link = 'logit'), na.action = "na.fail", nAGQ = 0)
```

```{r}
msummary(model_logit_fix_attempt)
```

Well, at least it's not giving NA's like with glmmTMB.

```{r}
acf(resid(model_logit_fix_attempt))
```

And that didn't do anything.  Time to ask Prof. DeRuiter for help.

Maybe a logistic regression model with a binary response isn't suitable in this scenario?

Continuing on for kicks and giggles though.

___

DHARMa package scaled residuals versus fitted (predicted) plot:

```{r}
model_logit_sim <- simulateResiduals(model_logit, n = 1000)
```

```{r}
gf_point(model_logit_sim$scaledResiduals ~ fitted(model_logit)) %>%
  gf_labs(x = "Predicted Probability for 'yes' is a drug-related",
          y = 'Scaled Residuals')
```

Mean-Variance relationship does seem okay as this looks like a uniform distribution.  Linearity should be fine too from a glance though the area in the lower middle is a little sparse compared to the rest.

___

#### Plot each quantitative predictor binned against the response.

Code in the section below just checks that we are binning correctly and does some other prep in preparation for the linearity check plots.

```{r}
all_data_binned <- all_data %>%
  mutate(binned_deaths = cut_number(deaths, 10)) 
```

```{r}
all_data_binned_levels <- unique(all_data_binned$binned_deaths)
all_data_binned_levels
all_data_binned_observations = c()
```

```{r}
for (element in all_data_binned_levels) {
  print(element)
  value <- nrow(filter(all_data_binned, binned_deaths == element))
  print(value)
  all_data_binned_observations <- append(all_data_binned_observations, value)
}
print(all_data_binned_observations)
all_data_binned_observations <- c()
```

Bins look pretty good in terms of the number of available observations in each bin.

```{r}
median(all_data$deaths)
class(median(all_data$deaths))
```

```{r}
prop(all_data$drug_death == "yes")
class(prop(all_data$drug_death == "yes"))
```

```{r}
glimpse(all_data_binned)
```

___

```{r}
all_data_grouped <- all_data %>%
  mutate(binned_deaths = cut_number(deaths, 10)) %>%
  group_by(binned_deaths) %>%
  summarize(proportion_drug_deaths = prop(drug_death == "yes"),
            median_deaths = median(deaths))
```

```{r}
head(all_data_grouped, 5)
```

```{r}
gf_point(proportion_drug_deaths ~ median_deaths, data = all_data_grouped,
         alpha = 1.0, color = "black",
         ylab="Proportion where drug_death == 'yes'",
         xlab="Median of 'deaths' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```

```{r}
gf_point(logit(proportion_drug_deaths) ~ median_deaths, data = all_data_grouped,
         alpha = 1.0, color = "black",
          ylab="Proportion where drug_death == 'yes'",
         xlab="Median of 'deaths' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```

Hmm, not quite a linear relationship.

___

```{r}
all_data_binned <- all_data %>%
  mutate(binned_population = cut_number(population, 10)) 
```

```{r}
all_data_binned_levels <- unique(all_data_binned$binned_population)
all_data_binned_levels
all_data_binned_observations = c()
```

```{r}
for (element in all_data_binned_levels) {
  print(element)
  value <- nrow(filter(all_data_binned, binned_population == element))
  print(value)
  all_data_binned_observations <- append(all_data_binned_observations, value)
}
print(all_data_binned_observations)
all_data_binned_observations <- c()
```

Bins look pretty good in terms of the number of available observations in each bin.

___

```{r}
all_data_grouped <- all_data %>%
  mutate(binned_population = cut_number(population, 10)) %>%
  group_by(binned_population) %>%
  summarize(proportion_drug_deaths = prop(drug_death == "yes"),
            median_population = median(population))
```

```{r}
head(all_data_grouped, 5)
```

```{r}
gf_point(proportion_drug_deaths ~ median_population, data = all_data_grouped,
         alpha = 1.0, color = "black",
         ylab="Proportion where drug_death == 'yes'",
         xlab="Median of 'population' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```

```{r}
gf_point(logit(proportion_drug_deaths) ~ median_population, data = all_data_grouped,
         alpha = 1.0, color = "black",
         ylab="Logit(Proportion) where drug_death == 'yes'",
         xlab="Median of 'population' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```

Yea...

___

```{r}
all_data_binned <- all_data %>%
  mutate(binned_percent_of_total_deaths = cut_number(percent_of_total_deaths, 10)) 
```

```{r}
all_data_binned_levels <- unique(all_data_binned$binned_percent_of_total_deaths)
all_data_binned_levels
all_data_binned_observations = c()
```

```{r}
for (element in all_data_binned_levels) {
  print(element)
  value <- nrow(filter(all_data_binned, binned_percent_of_total_deaths == element))
  print(value)
  all_data_binned_observations <- append(all_data_binned_observations, value)
}
print(all_data_binned_observations)
all_data_binned_observations <- c()
```

Bins look pretty good in terms of the number of available observations in each bin.

___

```{r}
all_data_grouped <- all_data %>%
  mutate(binned_percent_of_total_deaths = cut_number(percent_of_total_deaths, 10)) %>%
  group_by(binned_percent_of_total_deaths) %>%
  summarize(proportion_drug_deaths = prop(drug_death == "yes"),
            median_percent_of_total_deaths = median(percent_of_total_deaths))
```

```{r}
head(all_data_grouped, 5)
```

```{r}
gf_point(proportion_drug_deaths ~ median_percent_of_total_deaths, data = all_data_grouped,
         alpha = 1.0, color = "black",
         ylab="Proportion where drug_death == 'yes'",
         xlab="Median of 'percent_of_total_deaths' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```

```{r}
gf_point(logit(proportion_drug_deaths) ~ median_percent_of_total_deaths, data = all_data_grouped,
         alpha = 1.0, color = "black",
         ylab="Logit(Proportion) where drug_death == 'yes'",
         xlab="Median of 'percent_of_total_deaths' in each Bin",
         title="CDC Wonder Underlying Cause of Death Dataset Queries",
         subtitle = "",
         caption = ""
) %>% gf_lm()
```


Hmm, just noticed that you might be able to drive percent_of_total_deaths by deaths / population similarly to how crude_rate is deaths / population * 1000.

Will look into it at some point.

___

# Prediction plots (for the 3rd research question).

Get the fixed values we will be using for the plots.

```{r}
ggpredict(model_logit)
```

```{r}
t(get_fixed(model_logit)) %>% pander()
```

```{r}
get_fixed(model_logit)
```

```{r}
glimpse(all_data)
```

And our model is obviously terrible.

Prediction plot for "gender":

```{r}
ggpredict(model_logit, 
          terms = 'gender',
          type = 'fixed') %>% plot()
```

Prediction plot for "race":

```{r}
ggpredict(model_logit, 
          terms = 'race',
          type = 'fixed') %>% plot()
```

Prediction plot for "year":

```{r}
ggpredict(model_logit, 
          terms = 'year',
          type = 'fixed') %>% plot()
```

Prediction plot for "ten_year_age_groups":

```{r}
ggpredict(model_logit, 
          terms = 'ten_year_age_groups',
          type = 'fixed') %>% plot()
```

Prediction plot for "deaths":

```{r}
ggpredict(model_logit, 
          terms = 'deaths',
          type = 'fixed') %>% plot()
```

Prediction plot for "population":

```{r}
ggpredict(model_logit, 
          terms = 'population',
          type = 'fixed') %>% plot()
```

Prediction plot for "scaled_percent_of_total_deaths":

```{r}
ggpredict(model_logit, 
          terms = 'percent_of_total_deaths',
          type = 'fixed') %>% plot()
```

Prediction plot for "gender*race":

```{r}
ggpredict(model_logit, 
          terms = c('gender', 'race'),
          type = 'fixed') %>% plot()
```

___
___
___










































