---
title: "CSR Naloxone Standing Order Project - Draft Model(s)"
author: 'STAT 245, Fall 2020 (Group Members: Nana Ama Baidoo, Alex Visser, Joshua Ridder, Joseph Jinn)'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    fig_height: 8
    fig_width: 13
  html_document:
    fig_height: 8
    fig_width: 13
classoption: landscape
---

```{r, setup, include = FALSE}
# load packages that are going to be used
require(tidyverse)
require(ggformula)
require(mosaic)
require(fastR2)
require(s245)
require(pander)
require(DHARMa)
require(glmmTMB)
require(MuMIn)
require(car)
require(dplyr) # SQL syntax ftw.
require(readr)
# require(reticulate) # Utilize Python programming language.
require(ggeffects)

# Some customization.  You can alter or delete as desired (if you know what you are doing).

theme_set(theme_minimal(base_size=12))     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  echo = TRUE,      # for homework, always show R code (this is the default)
  error = TRUE, # do not interrupt generation in case of errors,
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small",   # slightly smaller font for code
  message = FALSE, warning = FALSE) # don't print warnings or messages in compiled document. So you MUST check them in RStudio!
```

Note to self: Run "remotes::install_github('ProjectMOSAIC/ggformula')" to install development version of ggformula.

```{r}
# sessionInfo()
options(max.print = 6000) # Ensure we can print entire summary.
# remotes::install_github('ProjectMOSAIC/ggformula')
```

# Data Tidying (yet again)

[readr documentation](https://www.rdocumentation.org/packages/readr/versions/1.3.1)

[read_table documentation](https://readr.tidyverse.org/reference/read_table.html)

[read_delim documentation](https://www.rdocumentation.org/packages/readr/versions/1.3.1/topics/read_delim)

[Data Tidying Tutorial](https://rsconnect.calvin.edu:3939/content/24/#section-other-data-file-formats)

```{r}
parse_query <- function(file_name_as_string) {
  data <- read_delim(file = file_name_as_string,
                     delim = "\t",
                     col_names = TRUE,
                     col_types = NULL,
                     na = "NA",
                     skip = 0,
                     n_max = Inf,
                     guess_max = 100,
                     progress = show_progress(),
                     comment =c("-"),
                     skip_empty_rows = TRUE
  )[, -c(1)]  %>% # Ignore the "Notes" column
    filter_all(any_vars(!is.na(.)))  # Remove all rows with ALL NAs in all columns (due to "Notes" section)
  return(data)
}
```

Parse the file(s).

```{r}
data_drug <- parse_query("datasets//joseph_jinn_queries//CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneDrugInducedCausesAllDates9decimalsPer1k.txt")
```

```{r}
glimpse(data_drug)
```

```{r}
data_alcohol <- parse_query("datasets//joseph_jinn_queries//CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneAlcoholInducedCausesAllDates9decimalsPer1k.txt")
```

```{r}
glimpse(data_alcohol)
```

Ignoring the errors. Error messages seem to indicate failing to parse the "Notes" section which we do not care about as they aren't data.

```{r}
data_other <- parse_query("datasets//joseph_jinn_queries//CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneAllOtherNonDrugNonAlcoholCausesAllDates9decimalsPer1k.txt")
```

```{r}
glimpse(data_other)
```

___

Add new variable specifying whether it was "alcohol", "drug", or "other" type of cause for death.

```{r}
data_drug <- data_drug %>% mutate(`Cause Of Death` = "drug")
data_alcohol <- data_alcohol %>% mutate(`Cause Of Death` = "alcohol")
data_other <- data_other %>% mutate(`Cause Of Death` = "other")
```


```{r}
glimpse(data_drug)
glimpse(data_alcohol)
glimpse(data_other)
```

___

Convert logical gender code to character gender code (R arbitrarily set "Female" = FALSE and "Male" = NA.

```{r}
head(data_drug, 5)
```

```{r}
tail(data_drug, 5)
```

___

```{r}
data_drug <- data_drug %>% mutate(`Gender Code` = case_when(`Gender` ==  'Female' ~ 'F',
                                                            TRUE ~ 'M'))
```

```{r}
head(data_drug, 5)
```

```{r}
tail(data_drug, 5)
```

```{r}
data_other <- data_other %>% mutate(`Gender Code` = case_when(`Gender` ==  'Female' ~ 'F',
                                                              TRUE ~ 'M'))
```

```{r}
head(data_other, 5)
```

```{r}
tail(data_other, 5)
```

___

[DPLYR joins](https://dplyr.tidyverse.org/reference/join.html)

[Join explanations using DPLYR](https://statisticsglobe.com/r-dplyr-join-inner-left-right-full-semi-anti)

Join the drug, alcohol, and other data for a complete dataset.

```{r}
data <- bind_rows(data_drug, data_alcohol, data_other)
```

```{r}
glimpse(data)
```

Ok, looking good so far.

___

[CDC Wonder Data Queries](https://wonder.cdc.gov/ucd-icd10.html)

[Other Data](https://www.cdc.gov/nchs/nvss/vsrr/drug-overdose-data.htm)

[Other Data](https://www.cdc.gov/drugoverdose/data/statedeaths/drug-overdose-death-2018.html)

[Basic Query Template](https://wonder.cdc.gov/controller/saved/D76/D95F739)

* Common shared columns (if included in result set):
+ Deaths
+ Population
+ `Crude Rate`
+ `Crude Rate Lower 95% Confidence Interval`
+ `Crude Rate Upper 95% Confidence Interval`
+ `Crude Rate Standard Error`
+ `% of Total Deaths`

Unique columns are based on what we group by (as seen in the dataframe above).

We should also rename columns to not have spaces and other special characters for ease with working with ggformula and other packages (so names aren't enclosed in "`").

Define function to replace "messy" column names with "clean" version.  Adjust in the future as necessary.

[R - Regular Expressions](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf)

```{r}
remove_whitespace_column_names <- function(my_dataframe) {
  
  return(names(my_dataframe) %>% str_replace_all("\\s", "_") %>% 
           str_replace_all("%", "percent") %>% 
           str_replace_all("/", "_") %>% 
           str_replace_all("-", "_") %>% 
           str_replace_all("^([0-9]+)", "c\\1") %>% # Replace beginning number with dummy c - character.
           tolower)
}
```

Clean/tidy the column names:

```{r}
colnames(data) <- remove_whitespace_column_names(data)
```

Check that "%", "-", and "/" have been removed, in addition to whitespace.

```{r}
glimpse(data)
```

___

If necessary:

Replace all rows with "Suppressed" in the "Deaths" column with randomly generated values between 0-9 since that is the criteria for when data values are labeled as "Suppressed".  

Replace all rows with "Suppressed" in the "Population" column with randomly generated values between 0-9 since that is the criteria for when data values are labeled as "Suppressed".

Crude rates are marked as "Unreliable" when death counts are less than 20.  We could manually create a crude rate by dividing "population" by "deaths" multiplied by X individuals.

Crude Rate = Count / Population * (per specified # of people)

Crude rates are marked as "Not Applicable" when "Population" is unavailable.

```{r}
count(data$deaths == "Suppressed")
count(data$population == "Suppressed")
count(data$crude_rate == "Suppressed")
count(data$crude_rate == "Unreliable")
count(data$crude_rate == "Not Applicable")
```

Good, we have no "Suppressed" rows but we do have a decent amount of "Unreliable" rows for crude rates.

Since the crude rates for our dataset were calculated per 1,000 people, we will use that for manual creation of crude rates for "Unreliable" rows.

**TODO: Compute crude rates for "Unreliable" rows and insert into dataset**

```{r}
data_temp <- data %>% mutate(crude_rate_added = case_when(
  crude_rate == 'Unreliable' ~ deaths / population * 1000,
  TRUE ~ as.numeric(crude_rate)
))
```

```{r}
glimpse(data_temp)
```

And double-check that our manual computations for crude_rate makes sense.

```{r}
# which(data$crude_rate == "Unreliable")
```

Row 9 and 12, among many others had "Unreliable" as the crude_rate.  With our manual computations of crude_rate, we seem to have obtained reasonable values.

Also, we have copied over crude_rate values for rows that had existing values for crude_rate successfully.  So, we are all set.

```{r}
head(data, 20)
```
```{r}
head(data_temp, 20)
```

And re-asssign temporary dataframe back to our original dataframe.

```{r}
data <- data_temp
```
___

Now, we should convert the data types for each attribute to the ones desired for model fitting.

```{r}
data <- data %>% mutate_if(is.character, as.factor)
```

```{r}
glimpse(data)
```

We will keep crude rate as a "factor" for now.  It does contain "Unreliable" for quite a few rows.

```{r}

```

And county_code, year, and year_code, should be a factor, not a numeric (double) type.

```{r}
data <- data %>% mutate(county_code = as.factor(county_code))
data <- data %>% mutate(year = as.factor(year))
data <- data %>% mutate(year_code = as.factor(year_code))
```

```{r}
glimpse(data)
```

And we want percent_of_total_deaths as a numeric type too not a factor or character with a % sign.

```{r}
data <- data %>% mutate(percent_of_total_deaths = as.numeric(gsub("%", "", percent_of_total_deaths)))
```

```{r}
glimpse(data)
```

And last but not least, "deaths" and "population" should be integer instead of double numeric sub-types.

```{r}
data <- data %>% mutate(deaths = as.integer(deaths))
data <- data %>% mutate(population = as.integer(population))
```

```{r}
glimpse(data)
```

Ok, things are looking good now.

But, we should also export out our cleaned and combined dataset to a CSV file for future use.

```{r}
# write.csv2(data,"datasets\\joseph_jinn_queries\\CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneAllCausesOfDeathDates9decimalsPer1k_TIDY.csv", append = FALSE, row.names = FALSE)
```

And test we can read everything back in correctly.

```{r}
import_test <- read_csv2("datasets\\joseph_jinn_queries\\CDCWonderDataQuery_UnderlyingCauseOfDeath_GROUPBY_GenderRaceYearTenYearAgeGroupCounty_LIMITBY_noneAllCausesOfDeathDates9decimalsPer1k_TIDY.csv")
```

```{r}
glimpse(import_test)
```

Well, I haven't figured out a way to store column data types when exporting to CSV (if that's even possible).  So, we will need to reconvert to the correct data types for each column.

```{r}
import_test <- import_test %>% mutate_if(is.character, as.factor)
```

```{r}
import_test <- import_test %>% mutate(county_code = as.factor(county_code))
import_test <- import_test %>% mutate(year = as.factor(year))
import_test <- import_test %>% mutate(year_code = as.factor(year_code))
```

```{r}
import_test <- import_test %>% mutate(percent_of_total_deaths = as.numeric(gsub("%", "", percent_of_total_deaths)))
```

```{r}
import_test <- import_test %>% mutate(deaths = as.integer(deaths))
import_test <- import_test %>% mutate(population = as.integer(population))
```

```{r}
glimpse(import_test)
```

```{r}
head(import_test)
```

And everything looks good.

___
